A Wait-free Queue with Poly-logarithmic Step Complexity

### 2022-12-17 podc_paper/introduction.tex
working on related work

There has been a great deal of research in the past several decades on the design of linearizable, lock-free queues.
Besides being a fundamental data structure, queues are used in
significant concurrent applications, including OS kernels \cite{MP91}, memory reclamation \cite{?},
packet processing \cite{DPDK},
and sharing resources or tasks.
The lock-free MS-queue of Michael and Scott \cite{MS98} is a classic shared queue implementation.
IT uses a singly-linked list with pointers to the front and back nodes.
To dequeue or enqueue an element, the front or back pointer is updated by a 
compare-and-swap (CAS) instruction.
If this CAS fails, the operation must retry.
In the worst case, this means that each successful CAS may cause all other processes to
fail and retry, leading to an amortized step complexity of $\Omega(p)$ per operation.
(To measure amortized step complexity of a lock-free implementation, we consider all possible finite executions
and divide the number of steps in the execution by the number of operations initated in the execution.)
Numerous papers have suggested modifications to the MS-queue \cite{DBLP:conf/opodis/HoffmanSS07,DBLP:conf/podc/KoganH14,DBLP:conf/ppopp/KoganP11,DBLP:journals/dc/Ladan-MozesS08,MKLLP22,DBLP:conf/spaa/MoirNSS05,RC17}, but 
all still have $\Omega(p)$ amortized step complexity as a result of
contention on the front and back of the queue.
Morrison and Afek \cite{DBLP:conf/ppopp/MorrisonA13} called this the CAS retry problem.
The same problem occurs in array-based implementations of queues \cite{DBLP:conf/iceccs/ColvinG05,DBLP:conf/icdcn/Shafiei09,DBLP:conf/spaa/TsigasZ01,DBLP:conf/opodis/GidenstamST10}.
Solutions that tried to sidestep this problem using fetch-and-increment \cite{DBLP:conf/ppopp/MorrisonA13,DBLP:conf/ppopp/YangM16,Nik19,10.1145/3490148.3538572}
rely on slower mechanisms to handle worst-case executions and still have $\Omega(p)$ step complexity.

@@

In this paper, we exploit this loophole.  We show that it is in fact possible to implement a linearizable, lock-free queue
whose step complexity does not depend linearly on $p$.
Our implementation is the first to have step complexity that is polylogarithmic in $p$ and in $q$, the number of elements
stored in the queue.
For simplicity of presentation, we first give an unbounded-space construction that uses
$O(\log^2 p + \log q)$ steps per operation.
Then, we show how to modify the implementation to use 
\here{$O(?)$ space} and \here{$O(?)$ steps per operation}.
Both implementations use single-word CAS instructions and reasonably-sized words.
Moreover, they use $O(\log p)$ CAS instructions per operation in the worst case, whereas previous
lock-free queues use an unbounded number of CAS instructions in the worst case
and $\Omega(p)$ CAS instructions in an amortized sense.
For the space-bounded implementation, we unlink unneeded memory from our data structure.
We do not address the orthogonal problem of reclaiming unlinked memory; we assume a safe
garbage collector, such as the highly optimized one Java provides.

>>>>

There has been a great deal of research in the past several decades on the design of linearizable, lock-free queues.
Besides being a fundamental data structure, queues are used in
significant concurrent applications, including OS kernels \cite{MP91}, memory reclamation \cite{?},
packet processing \cite{DPDK},
and sharing resources or tasks.
The lock-free MS-queue of Michael and Scott \cite{MS98} is a classic shared queue implementation.
IT uses a singly-linked list with pointers to the front and back nodes.
To dequeue or enqueue an element, the front or back pointer is updated by a 
compare-and-swap (CAS) instruction.
If this CAS fails, the operation must retry.
In the worst case, this means that each successful CAS may cause all other processes to
fail and retry, leading to an amortized step complexity of $\Omega(p)$ per operation.
(To measure amortized step complexity of a lock-free implementation, we consider all possible finite executions
and divide the number of steps in the execution by the number of operations initated in the execution.)
Numerous papers have suggested modifications to the MS-queue \cite{DBLP:conf/opodis/HoffmanSS07,DBLP:conf/podc/KoganH14,DBLP:conf/ppopp/KoganP11,DBLP:journals/dc/Ladan-MozesS08,MKLLP22,DBLP:conf/spaa/MoirNSS05,RC17}, but 
all still have $\Omega(p)$ amortized step complexity as a result of
contention on the front and back of the queue.
Morrison and Afek \cite{DBLP:conf/ppopp/MorrisonA13} called this the CAS retry problem.
The same problem occurs in array-based implementations of queues \cite{DBLP:conf/iceccs/ColvinG05,DBLP:conf/icdcn/Shafiei09,DBLP:conf/spaa/TsigasZ01,DBLP:conf/opodis/GidenstamST10}.
Solutions that tried to sidestep this problem using fetch\&increment \cite{DBLP:conf/ppopp/MorrisonA13,DBLP:conf/ppopp/YangM16,Nik19,10.1145/3490148.3538572}
rely on slower mechanisms to handle worst-case executions and still have $\Omega(p)$ step complexity.

@@

In this paper, we exploit this loophole.  We show that it is in fact possible to implement a linearizable, lock-free queue
whose step complexity does not depend linearly on $p$.
Our implementation is the first to have step complexity that is polylogarithmic in $p$ and in $q$, the number of elements
stored in the queue.
For simplicity of presentation, we first give an unbounded-space construction that uses
$O(\log^2 p + \log q)$ steps per operation.
Then, we show how to modify the implementation to use 
\here{$O(?)$ space} and \here{$O(?)$ steps per operation}.
Both implementations use single-word CAS instructions and reasonably-sized words.
Moreover, they use $O(\log p)$ CAS instructions per operation in the worst case, whereas previous
lock-free queues use 
%an unbounded number of CAS instructions in the worst case and 
$\Omega(p)$ CAS instructions in an amortized sense.
For the space-bounded implementation, we unlink unneeded memory from our data structure.
We do not address the orthogonal problem of reclaiming unlinked memory; we assume a safe
garbage collector, such as the highly optimized one that Java provides.

### 2022-12-17 podc_paper/related.tex
working on related work

Morrison and Afek \cite{DBLP:conf/ppopp/MorrisonA13} also used a linked list of (circular) arrays.
To avoid the CAS retry problem, concurrent operations try to claim spots in an array using fetch-and-increment instructions.
If livelock between enqueues and a racing dequeue prevent enqueues from claiming a spot,
the enqueues fall back on using a CAS to add a new array to the linked list, 
and the CAS retry problem reappears.
% I looked at the CRQ algorithm to implement each of the circular arrays.
% Here is an execution that causes livelock for 3 processes
% enqueue(A) reads t=0 and increments tail to 1
% enqueue(B) reads t=1 and increments tail to 2
% (*) dequeue reads h=0 and increments head to 1
% dequeue's CAS succeds changing array[0] to (1,R,bottom)
% dequeue reads t=2 at line 53
% enqueue(A) fails CAS on array[0]
% dequeue reads h=1 and increments head to 2
% dequeue's CAS succeeds changing array[1] to (1,R+1,bottom)
% enqueue(B) fails CAS on array[1]
% enqueue(A) reads t=2 and increments tail to 3
% enqueue(B) reads t=3 and increments tail to 4
% dequeue reads t=4 at line 53
% go back to (*) and repeat this sequence
Similarly,
other queues \cite{Nik19,10.1145/3490148.3538572,DBLP:conf/ppopp/YangM16} used the fast-path slow-path methodology of Kogan and Petrank~\cite{10.1145/2370036.2145835}.
The fast path has good performance and the slow path guarantees
termination.
In worst-case executions, where processes must execute the slow path,
these implementations also take $\Omega(p)$ steps per operation, 
due either to the CAS retry problem or helping mechanisms.

@@

\paragraph{Universal Constructions and Other Poly-log Time Data Structures}
A \textit{universal construction} is an algorithm that can implement a
shared version of any given sequential object. The first universal
construction was introduced by
Herlihy~\cite{10.1145/114005.102808}. We can implement a concurrent
queue using a universal construction. Jayanti proved an $\Omega(\log
p)$ lower bound on the worst-case shared-access time complexity of
$p$-process universal
constructions~\cite{DBLP:conf/podc/Jayanti98a}. He also mentions that
the universal construction by Afek, Dauber, and
Touitou~\cite{DBLP:conf/stoc/AfekDT95} can be modified to $O(\log p)$
worst-case step complexity, using atomic access to $\Omega(p \log
p)$-bit words. Chandra, Jayanti and Tan introduced a semi-universal
construction that achieves $\textsc{O}(\log^2 p)$ shared
accesses~\cite{DBLP:conf/podc/ChandraJT98}. However, their algorithm
cannot be used to create a queue. We mention a non-practical universal
construction with a poly-log number of \nf{CAS} instructions in the
last paragraph of page 13.

Need to discuss Fatourou and Kallimanis TCS 55(3)
Fatourou and Kallimanis \cite{FK14} gave a universal construction based on fetch-and-add and CAS
and used it to implement a queue....

Ellen and Woelfel introduced an implementation of a Fetch\&Inc object
with step complexity of $O(\log p)$ using $O(\log n$)-bit
\texttt{LL/SC} objects, where $n$ is the number of
operations~\cite{10.1007/978-3-642-41527-2_20}. Their idea to achieve
logarithmic complexity is to use a tree storing the Fetch\&Inc
operations invoked by processes. When a process wants to do a
Fetch\&Inc it adds its Fetch\&Inc to the tree and returns the number
of elements in the tree. There are some similarities between designing
a queue and a Fetch\&Inc object. A Fetch\&Inc object can be
constructed from a queue. The algorithm by Ellen and Woelfel is
interesting because of the similarities between Fetch\&Inc objects and
queues. Also, it is one of the few wait-free data structures achieving
poly-logarithmic complexity.

@@

>>>>

Morrison and Afek \cite{DBLP:conf/ppopp/MorrisonA13} also used a linked list of (circular) arrays.
To avoid the CAS retry problem, concurrent operations try to claim spots in an array using fetch\&increment instructions.
If livelock between enqueues and a racing dequeue prevent enqueues from claiming a spot,
the enqueues fall back on using a CAS to add a new array to the linked list, 
and the CAS retry problem reappears.
% I looked at the CRQ algorithm to implement each of the circular arrays.
% Here is an execution that causes livelock for 3 processes
% enqueue(A) reads t=0 and increments tail to 1
% enqueue(B) reads t=1 and increments tail to 2
% (*) dequeue reads h=0 and increments head to 1
% dequeue's CAS succeds changing array[0] to (1,R,bottom)
% dequeue reads t=2 at line 53
% enqueue(A) fails CAS on array[0]
% dequeue reads h=1 and increments head to 2
% dequeue's CAS succeeds changing array[1] to (1,R+1,bottom)
% enqueue(B) fails CAS on array[1]
% enqueue(A) reads t=2 and increments tail to 3
% enqueue(B) reads t=3 and increments tail to 4
% dequeue reads t=4 at line 53
% go back to (*) and repeat this sequence
Similarly,
other queues \cite{Nik19,10.1145/3490148.3538572,DBLP:conf/ppopp/YangM16} used the fast-path slow-path methodology of Kogan and Petrank~\cite{10.1145/2370036.2145835}.
The fast path has good performance and the slow path guarantees
termination.
In worst-case executions, where processes must execute the slow path,
these implementations also take $\Omega(p)$ steps per operation, 
due either to the CAS retry problem or helping mechanisms.

@@

\paragraph{Universal Constructions}
One can also build a queue using a universal construction \cite{10.1145/114005.102808}.
Jayanti \cite{DBLP:conf/podc/Jayanti98a} observed that
the universal construction of Afek, Dauber, and
Touitou~\cite{DBLP:conf/stoc/AfekDT95} can be modified to use $O(\log p)$ steps per operation, 
using atomic access to $\Omega(p \log p)$-bit words. 
(Thus, in terms of operations on reasonably-sized words, their construction would still take $\Omega(p)$ steps per operation.)
Fatourou and Kallimanis \cite{FK14} used their universal construction based on fetch\&add and LL/SC instructions
to implement a queue, but its step complexity is also $\Omega(P)$.
\here{Double check this.}

@@

\paragraph{Fetch\&Increment Objects}
Ellen and Woelfel introduced an implementation of a Fetch\&Increment object
with step complexity of $O(\log p)$ using $O(\log n$)-bit
\texttt{LL/SC} objects, where $n$ is the number of
operations~\cite{10.1007/978-3-642-41527-2_20}. Their idea to achieve
logarithmic complexity is to use a tree storing the Fetch\&Increment
operations invoked by processes. When a process wants to do a
Fetch\&Increment it adds its Fetch\&Increment to the tree and returns the number
of elements in the tree. There are some similarities between designing
a queue and a Fetch\&Increment object. A Fetch\&Increment object can be
constructed from a queue. The algorithm by Ellen and Woelfel is
interesting because of the similarities between Fetch\&Increment objects and
queues. Also, it is one of the few wait-free data structures achieving
poly-logarithmic complexity.
